Finch: Sparse and Structured Array Programming with Control Flow

### 2024-04-05 sections/lang.tex
Update on Overleaf.
good till 5

\section{The Finch Language}

\begin{wrapfigure}{r}{0.51\linewidth}
    %\vspace{-26pt}
    \begin{minted}{julia}
     EXPR := LITERAL | VALUE | INDEX | VARIABLE | EXTENT | CALL | ACCESSS
     STMT := ASSIGN | LOOP | DEFINE | SIEVE | BLOCK

      DECLARE := TENSOR .= EXPR(EXPR...)   #V is the set of all values
       FREEZE := @freeze(TENSOR)           #S is the set of all Symbols
         THAW := @thaw(TENSOR)             #T is the set of all types                                      
       ASSIGN := ACCESS <<EXPR>>= EXPR    
         LOOP := for INDEX = EXPR              LITERAL := V                           
                   STMT                          VALUE := S::T                        
                 end                            TENSOR := S                            
       DEFINE := let VARIABLE = EXPR             INDEX := S                            
                   STMT                       VARIABLE := S                            
                 end                            EXTENT := EXPR : EXPR                  
        SIEVE := if EXPR                          CALL := EXPR(EXPR...)                
                   STMT                         ACCESS := TENSOR[EXPR...]              
                 end                              MODE := @mode(TENSOR)                
        BLOCK := begin                     
                   STMT...                 
                 end                        
    \end{minted}                                                   
    \vspace{-12pt}
\caption{The syntax of the finch language. Compare this grammar to the Concrete
Index Notation of TACO \cite[Figure~3]{kjolstad_tensor_2019}, noting the
addition of multiple left hand sides through code blocks, access with expressions, and explicit initialiation.}\label{fig:syntax}

\end{wrapfigure}

@@

Our syntax is highly permissive: by allowing blocks of code with multiple statements, we support both multiple outputs and temporary tensors, seen most prominently in our implementation of gustavson's algorithm for sparse-sparse matrix multiply and breadth-first-search.
%
Our breadth-first seach also uses an
\mintinline{julia}{if} statement to avoid operating on vertices outside the
frontier. 
%
We have found that \mintinline{julia}{if} is particularly useful in
our model for masking to avoid work.
%
However, to compile these constructs efficiently, we must restrict this syntax in several ways, as we have alluded to in the previous section, via life cycles.
%
In the semantics section, we detail other specifics that allow us to specialize our syntax efficient code over structured data.

@@

We include the sketch of our semantics to highlight several key design choices.
%
First, we note that the life cycle rules (in combination with the definition rules) prevent programs that read and write to values within a loop except in the narrow circumstance of updates, which cannot even be applied to scalar definitions.
%
Beyond the simple management of iterator state, these rules allow us to effectively analyze the propagation of constants within a loop.
%
We eliminate problems that in a less restrictive language would require alias analysis so that we can tell when a zero or repetitive value read from one tensor is being used in an expression or written to another tensor.
%
These restrictions play a similar role to that of a pure function in Halide~\cite{ragan-kelley_halide_2013}.
%
Second, loops with tensor accesses enter into a looplet system via the $Unfurl$ rule.
%
In this system, repeated structures and constants are slowly uncovered as accesses are lowered in various points in the program (e.g. $Run$  and $Switch$, respectively).
%
In this process, we are able to use rewrite rules in $Simplify$ to eliminate cases, unnecessary iterations, and so forth based on the information provided via Looplets and via the control flow (loops, sieve, definitions).
%

Though the reader may doubt the coverage of this approach, the next section will detail how the compiler adds additional features on top of this core language ultimately through the assurance that accesses to tensor values will first unravel to Looplets and then simplify in the context of the location of the access.
%
In the next section, we will handle complex index access expressions involving indirection and arithmetic by transforming complex accesses to additional loop variables with additional tensor accesses to insert Looplets to provide the structure dictated by the complex control flow.
%
Also in the next section, we will show how similar means lets us exploit the structure of complex conditional expressions when the structure of the expressions is placed inside a tensor access.

@@

>>>>

\vspace{12pt}
\section{The Finch Language}

\begin{wrapfigure}{R}{0.51\linewidth}
    \vspace{-28pt}
    \begin{minted}{julia}
     EXPR := LITERAL | VALUE | INDEX | VARIABLE | EXTENT | CALL | ACCESSS
     STMT := ASSIGN | LOOP | DEFINE | SIEVE | BLOCK

      DECLARE := TENSOR .= EXPR(EXPR...)   #V is the set of all values
       FREEZE := @freeze(TENSOR)           #S is the set of all Symbols
         THAW := @thaw(TENSOR)             #T is the set of all types                                      
       ASSIGN := ACCESS <<EXPR>>= EXPR    
         LOOP := for INDEX = EXPR              LITERAL := V                           
                   STMT                          VALUE := S::T                        
                 end                            TENSOR := S                            
       DEFINE := let VARIABLE = EXPR             INDEX := S                            
                   STMT                       VARIABLE := S                            
                 end                            EXTENT := EXPR : EXPR                  
        SIEVE := if EXPR                          CALL := EXPR(EXPR...)                
                   STMT                         ACCESS := TENSOR[EXPR...]              
                 end                              MODE := @mode(TENSOR)                
        BLOCK := begin                     
                   STMT...                 
                 end                        
    \end{minted}                                                 
    \vspace{-12pt}
\caption{The syntax of the finch language. Compare this grammar to the Concrete
Index Notation of TACO \cite[Figure~3]{kjolstad_tensor_2019}, noting the
addition of multiple left hand sides through code blocks, access with arbitrary expressions, and explicit declaration, as well as freeze and thaw.}\label{fig:syntax}
\vspace{-12pt}
\end{wrapfigure}

@@

Our syntax is highly permissive: by allowing blocks of code with multiple statements, we implicitly support many features gained through complicated scheduling commands in other frameworks, such as multiple outputs, masking to avoid work, temporary tensors, and arbitrary loop fusion and nesting.
%
These features are seen most prominently in our implementation of gustavson's algorithm for sparse-sparse matrix multiply, which simply writes to a temporary tensor in an inner loop and then reuses it, or in our breadth-first-search, which uses an \mintinline{julia}{if} statement to avoid operating on vertices outside the frontier. 
%
The only restriction we impose on our syntax is that it must respect tensor life cycles.
%
In the semantics section, we detail the specifics of how we compile our syntax to efficient code over structured data.

@@

We include the sketch of our semantics to highlight several key design choices.
%
First, we note that the life cycle rules (in combination with the definition rules) prevent programs that access tensors on both the left and the right hand side of an incrementing assignment within a loop.
%
Tensors may only change their mode in the scope in which they were declared.
%
The $\finchloop$, $\finchdeclare$, and $\finchsieve$ nodes introduce new scopes, and tensors without declarations
are assumed to have global scope.
%
Beyond the simple management of iterator state, these rules allow us to effectively analyze the propagation of constants, zeros, and repetitive values within a loop, eliminating problems that in a less restrictive language would require alias analysis.
%(similar to the role of a pure functions in Halide~\cite{ragan-kelley_halide_2013}).
%
%We eliminate problems that in a less restrictive language would require alias analysis to, for example, propagate a zero or repetitive value in a tensor.
%
Second, loops enter into a looplet system via the $Unfurl$ rule.
%
In this system, repeated structures and constants are slowly uncovered as accesses are lowered in various points in the program (e.g. $Run$  and $Switch$, respectively).
%
In this process, we are able to use rewrite rules in $Simplify$ to eliminate cases, unnecessary iterations, and so forth based on the information provided via Looplets and via the control flow (loops, sieve, definitions).
%

The next section will detail how the compiler adds additional features on top of this core language.  
We rely on the assurance that accesses to tensor values will first unravel to Looplets and then simplify in the context of the tensor access.
%
We will handle complex index access expressions involving indirection and arithmetic by transforming complex accesses to additional loop variables with additional tensor accesses to insert Looplets to provide the structure dictated by the complex control flow.
%
We will also show how we can similarly exploit the structure of complex conditional expressions when the structure of the expressions is placed inside a tensor access.

@@

\begin{figure}[b]
    \centering
    \footnotesize
